### 2013-07-23 doc/dissertation/document/chapters/introduction/introduction.tex
working on refactoring the introduction chapter

\chapter{Introduction}
\label{ch:introduction}
In nearly all high-fidelity nuclear reactor simulations, linear and
nonlinear transport problems are a primary focus of study. Neutronics,
fluid flow, and heat transfer are among the predominate physics that
fall into this category and simulation technologies to solve these
problems make up much of the analysis suite used for modern
analysis. Recent focus on multiple physics systems in the nuclear
reactor modeling and simulation community adds a new level of
complexity to common linear and nonlinear systems as solution
strategies change when they are coupled to other problems
\citep{u.s._department_of_energy_casl_2011}. Furthermore, a desire for
predictive simulations to enhance the safety and performance of
nuclear systems creates a need for extremely high fidelity
computations to be performed for these transport systems as a means to
capture effects not modeled by coarser methods.

In order to achieve this high fidelity, state-of-the-art computing
\index{facilities} must be leveraged in a way that is both efficient
and considerate of hardware-related issues. As scientific computing
moves beyond petascale facilities with machines of $O(1,000,000)$
cores already coming on-line, new algorithms to solve these complex
problems must be developed to leverage this new hardware
\citep{kogge_using_2011}. Issues such as resiliency to node failure,
limited growth of memory available per node, and scaling to large
numbers of cores will be pertinent to robust algorithms aimed at this
new hardware. Considering these issues, this dissertation develops a
massively parallel Monte Carlo method for linear problems and a novel
Monte Carlo method to advance solution techniques for nonlinear
problems.

This goal of this work is to research and develop parallel solutions
for discrete linear and nonlinear transport problems that improve
performance and scalabilty using a novel set of Monte Carlo synthetic
acceleration methods.

We discuss in this chapter physics-based motivation for advancing
Monte Carlo synthetic acceleration methods and studying their
application to transport systems by providing problems of interest in
nuclear reactor analysis. Hardware-based motivations are also provided
by considering the impact of forthcoming computing architectures. In
addition, background on the current solver techniques for these
physics problems and a brief comparison to the proposed methods are
provided to further motivate this work. The organization of the
document is then reviewed.

@@

Although solution techniques in each of these individual categories
has advanced over the last few decades and in fact leveraged modern
algorithms and computer architectures, true predictive capability for
engineered systems can only be achieved through a coupled, multiple
physics analysis where the effects of feedback between physics are
modeled. For example, consider the safety analysis of a departure from
nucleate boiling scenario in the subchannel of a nuclear fuel
assembly.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/introduction/dnb_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Multiphysics dependency analysis of departure from
      nucleate boiling.} \textit{A neutronics solution is required to
      compute power generation in the fuel pins, fluid dynamics is
      required to characterize boiling and fluid temperature and
      density, heat transfer is required to compute the fuel and
      cladding temperature, and the nuclear data modified with the
      temperature and density data. Strong coupling among the
      variables creates strong nonlinearities.}}
  \label{fig:dnb_example}
\end{figure}
When this event occurs, heat transfer is greatly reduced between the
fuel and the coolant due to the vapor layer generated by boiling,
causing the fuel center-line temperature to rapidly rise. To
characterize this boiling phenomena and how it affects fuel failure we
must consider a neutronics analysis in order to compute power
generation in the fuel pins, fluid dynamics analysis to characterize
coolant boiling, temperature, and density, solid material heat
transfer to characterize fuel and cladding temperature and heat
transfer with the coolant, and nuclear data processing to characterize
how changing material temperatures and densities changes the cross
sections needed for the neutronics calculation. As shown in
Figure~\ref{fig:dnb_example}, many couplings are required among
individual physics components in order to accurately model this
situation with each physics generating and receiving many
responses. Those variables that are very tightly coupled, such as the
temperatures generated by the fluid dynamics and heat transfer
components, will have strong nonlinearities in their behavior and
would therefore benefit from fully consistent nonlinear solution
schemes instead of fixed-point type iterations between
physics\footnote{Fixed-point iterations between physics are commonly
  referred to as Picard iterations.}. Furthermore, the space and time
scales over which these effects occur will also vary greatly.

The computational resources required to solve such problems are
tremendous. Recent work in modeling coupled fluid flow and solid
material heat and mass transfer in a reactor subsystem, similar to the
same components of the departure from nucleate boiling example above,
was performed as part of analysis of the Department of Energy's
Consortium for Advanced Simulation of Light Water Reactors (CASL)
modeling and simulation hub. CASL used the Drekar multiphysics code
developed at Sandia National Laboratories
\citep{pawlowski_drekar_2012} for modeling goals in
grid-to-rod-fretting analysis and will use a similar coupled physics
structure for future departure from nucleate boiling analysis with
comparison to experimental data. Using Drekar, multiphysics
simulations have been performed with fully consistent methods for the
solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
leveraging $O(100,000)$ cores on leadership class machines. Neutronics
components to be implemented in CASL for multiphysics analysis, such
as the Exnihilo radiation transport suite developed at Oak Ridge
National Laboratory \citep{evans_denovo:_2010}, compute trillions of
unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
meshes and $O(100,000)$ cores as well. Given the large scale and
complexity of these problems, if we aim to advance multiphysics
solution techniques, then we are motivated to advance the solution of
complex and general nonlinear problems exploiting leadership class
levels of parallelism.

@@

In order to leverage MCSA as a solution technique for physics
problems, its current formulation requires the linear operator and all
preconditioners to be explicitly formed. Recent developments in the
Exnihilo neutronics package at Oak Ridge National Laboratory have
permitted generation of the $SP_N$ system of equations for detailed
neutronics models of fission reactor-based
systems\citep{evans_simpli_2013}. By fully forming these equations and
formulating them as a linear algebra problem instead of using the
explicit iterative methods of the past, we now have access to all of
the modern advancements in computational linear algebra including
Krylov solvers for asymmetric systems and algebraic preconditioning
methods. This leads us to then explore the applicability of our work
in discrete Monte Carlo methods for linear systems as a possible
solution method for the $SP_N$ equations. In addition, solving the
$SP_N$ equations in this way also breaks away from the $S_N$ forms of
parallelism where spatial parallelism is achieved by an efficient
parallel sweep, angular efficiency achieved by pipe-lining, and energy
parallelism achieved by decoupling the groups. With the $SP_N$
equations as a full matrix system, we now can parallelize the problem
as prescribed by the linear solver, which may be significantly more
scalable than current $S_N$ transport practices.

@@

@@

%%---------------------------------------------------------------------------%%
\section{Research Outline}
\label{sec:research_outline}
For some time, the particle transport community has been utilizing
Monte Carlo methods for the solution of transport problems
\citep{lewis_computational_1993}. The partial differential equation
(PDE) community has focused on various deterministic methods for
solutions to linear problems \citep{saad_iterative_2003,
  kelley_iterative_1995}. In between these two areas are a not widely
known group of Monte Carlo methods for solving sparse linear systems
\citep{forsythe_matrix_1950, hammersley_monte_1964,
  halton_sequential_1962, halton_sequential_1994}. In recent years,
these methods have been further developed for radiation transport
problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
\citep{evans_monte_2009, evans_monte_2012} but have yet to be applied
to more general sparse linear systems commonly generated by the
computational physics community. Compared to other methods in this
regime, MCSA offers three attractive qualities; (1) the linear problem
operator need not be symmetric or positive-definite, thereby reducing
preconditioning complexity, (2) the stochastic nature of the solution
method provides a natural solution to the issue of resiliency, and (3)
is amenable to parallelization using modern methods developed by the
transport community \citep{wagner_hybrid_2010}. The development of
MCSA as a general linear solver and the development of a parallel MCSA
method are new and unique features of this work, providing a
framework with which other issues such as resiliency may be addressed
in the future.

@@

The Forward-Automated Newton-MCSA (FANM) method is developed as new
nonlinear solution method. The key features of FANM are: full Jacobian
generation using modern Forward Automated Differentiation (FAD)
methods \citep{bartlett_automatic_2006}, and MCSA as the inner linear
solver. This method has several attractive properties. First, the
first-order approximation to the Jacobian used in JFNK type methods is
eliminated by generating the Jacobian explicitly with the model
equations through FAD. Second, the Jacobian need not be explicitly
formed by the user but is instead automated through FAD; this
eliminates the complexity of hand-coding derivatives and has also been
demonstrated to be more efficient computationally than evaluating
difference derivatives. Third, unlike GMRES, MCSA does not build a
subspace during iterations. Although the Jacobian must be explicitly
formed to use MCSA, for problems that take more than a few GMRES
iterations to converge the size of the Krylov subspace will grow
beyond that of the Jacobian. Finally, using MCSA for the linear solve
provides its benefits for preconditioning, potential resiliency, and
parallelism.

%%---------------------------------------------------------------------------%%
\section{Outline}

To present these new developments this document is arranged in the
following manner. First, in Chapter~\ref{ch:stochastic_methods}, the
fundamentals of the Monte Carlo method for linear systems are
presented. Using this background on Monte Carlo, synthetic
acceleration methods are then presented and analyzed using a simple
model transport problem. Next, in Chapter~\ref{ch:spn_equations}, the
Monte Carlo methods developed in Chapter~\ref{ch:stochastic_methods}
are applied to the neutron transport problem. Specifically, the $SP_N$
form of the Boltzmann neutron transport equation is developed and
solved using the Monte Carlo methods. Using a difficult fission
reactor fuel assembly criticality calculation to drive research and
development for these Monte Carlo methods, several important issues
regarding Monte Carlo solver applicability and performance will be
discussed along with solutions to these problems. Using these
solutions, the Monte Carlo methods are then verified against modern
solution techniques for the $SP_N$ equations in order verify
correctness for the fuel assembly criticality calculation. In
Chapter~\ref{ch:nonlinear_problem}, the Monte Carlo methods are
applied to the Navier-Stokes equations as a model nonlinear energy and
momentum transport system. Additional difficulties that arise when
using Monte Carlo methods in these types of systems will be presented
along with the subsequently developed solutions. In
Chapter~\ref{ch:parallel_methods}, the Monte Carlo methods are
parallelized using modern reactor physics techniques for particle
transport with a full parallel scaling analysis provided using a
leadership-class computing facility. Finally, the work is summarized
in Chapter~\ref{ch:conclusion} and topics of future work derived from
the results of this research presented.

>>>>

\chapter{Introduction}
\label{ch:introduction}
In nearly all high-fidelity nuclear reactor simulations, linear and
nonlinear transport problems are a primary focus of study. Neutronics,
fluid flow, and heat transfer are among the predominate physics that
fall into this category and simulation technologies to solve these
problems make up much of the analysis suite used for modern
engineering calculations. Recent focus on multiple physics systems in
the nuclear reactor modeling and simulation community adds a new level
of complexity to common linear and nonlinear systems as solution
strategies change when they are coupled to other problems
\citep{u.s._department_of_energy_casl_2011}. Furthermore, a desire for
predictive simulations to enhance the safety and performance of
nuclear systems creates a need for extremely high fidelity
computations to be performed for these transport systems as a means to
capture effects not modeled by coarser methods.

In order to achieve this high fidelity, state-of-the-art computing
\index{facilities} must be leveraged in a way that is both efficient
and considerate of hardware-related issues. As scientific computing
moves beyond petascale facilities with machines of $O(1,000,000)$
cores already coming on-line, new algorithms to solve these complex
problems must be developed to leverage this new hardware
\citep{kogge_using_2011}. Issues such as resiliency to node failure,
limited growth of memory available per node, and scaling to large
numbers of cores will be pertinent to robust algorithms aimed at this
new hardware. Considering these issues, this work researches and
develops parallel Monte Carlo Synthetic Acceleration methods for
linear problems and a new method based on the synthetic acceleration
techniques to advance solution algorithms for nonlinear problems.

We discuss in this chapter physics-based motivation for advancing
Monte Carlo Synthetic Acceleration methods and studying their
application to transport systems by providing problems of interest in
nuclear reactor analysis. Hardware-based motivations are also provided
by considering the potential impact of these methods on forthcoming
computing architectures. In addition, background on the current solver
techniques for these physics problems and a brief comparison to the
proposed methods are provided to further motivate this work. Finally,
the statement of work is given and the organization of the document is
then reviewed.

@@

Although solution techniques in each of these individual categories
has advanced over the last few decades and in fact leveraged modern
algorithms and computer architectures, true predictive capability for
engineered systems can only be achieved through a coupled, multiple
physics analysis where the effects of feedback between physics are
modeled. For example, consider the safety analysis of a departure from
nucleate boiling scenario in the subchannel of a nuclear fuel
assembly.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/introduction/dnb_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Multiphysics dependency analysis of departure from
      nucleate boiling.} \textit{A neutronics solution is required to
      compute power generation in the fuel pins, fluid dynamics is
      required to characterize boiling and fluid temperature and
      density, heat transfer is required to compute the fuel and
      cladding temperature, and the nuclear data modified with the
      temperature and density data. Strong coupling among the
      variables creates strong nonlinearities.}}
  \label{fig:dnb_example}
\end{figure}

When this event occurs, heat transfer is greatly reduced between the
fuel and the coolant due to the vapor layer generated by boiling,
causing the fuel center-line temperature to rapidly rise. To
characterize this boiling phenomena and how it affects fuel failure we
must consider a neutronics analysis in order to compute power
generation in the fuel pins, fluid dynamics analysis to characterize
coolant boiling, temperature, and density, solid material heat
transfer to characterize fuel and cladding temperature and heat
transfer with the coolant, and nuclear data processing to characterize
how changing material temperatures and densities changes the cross
sections needed for the neutronics calculation.

As shown in Figure~\ref{fig:dnb_example}, many couplings are required
among individual physics components in order to accurately model this
situation with each physics generating and receiving many
responses. Those variables that are very tightly coupled, such as the
temperatures generated by the fluid dynamics and heat transfer
components, will have strong nonlinearities in their behavior and
would therefore benefit from fully consistent nonlinear solution
schemes such as that developed by this work instead of fixed-point
type iterations between physics\footnote{Fixed-point iterations
  between physics are commonly referred to as Picard iterations.}. To
leverage such a method for analysis such as a departure from nucleate
boiling calculation, we also require effective solutions for each of
the physics involved including the neutronics problem and the fluid
problem, both of which we discuss next.

@@

In order to leverage Monte Carlo Synthetic Acceleration as a solution
technique for physics problems, its current formulation requires the
physics operator and all preconditioners to be explicitly
formed. Recent developments in the Exnihilo neutronics package at Oak
Ridge National Laboratory have permitted generation of the $SP_N$
system of equations for detailed neutronics models of fission
reactor-based systems\citep{evans_simpli_2013}. By fully forming these
equations and formulating them as a linear algebra problem instead of
using the explicit iterative methods of the past, we now have access
to all of the modern advancements in computational linear algebra
including Krylov solvers for asymmetric systems and algebraic
preconditioning methods. This leads us to then explore the
applicability of our work in discrete Monte Carlo methods for linear
systems as a possible solution method for the $SP_N$ equations. In
addition, solving the $SP_N$ equations in this way also breaks away
from the $S_N$ forms of parallelism where spatial parallelism is
achieved by an efficient parallel sweep, angular efficiency achieved
by pipe-lining, and energy parallelism achieved by decoupling the
groups. With the $SP_N$ equations as a full matrix system, we now can
parallelize the problem as prescribed by the linear solver, which may
be significantly more scalable than current $S_N$ transport practices.

@@

Nonlinear transport problems are a common occurrence in single and
multiple physics problems in nuclear engineering. Systems of partial
differential equations, such as the Navier-Stokes equations that
describe fluid flow, yield discrete sets of stiff equations with
nonlinearities present in the variables when discretized by
conventional methods. These sets of equations characterize momentum
and energy transport through diffusion, viscous forces and convection,
and inertial forces. Traditionally, such systems have been solved by
linearizing them in a form where the nonlinearities in the variables
are eliminated and more traditional linear methods can be used for
solutions. Often characterized as segregated methods where physics
operators are split and their action on the system approximated in
steps, such methods lack consistency and accuracy in resolving the
nonlinear component of the solution. In the last 30 years, fully
consistent nonlinear methods based on Newton's method have become more
popular and many advances have been made in the nuclear engineering
field to employ these methods.

In the context of solving standalone linear systems, Monte Carlo
methods do not provide significant merit over subspace methods due to
the fact that the physics operator must be explicitly formed. For many
applications including higher fidelity neutron transport
discretizations beyond the $SP_N$ equations discussed in the previous
section, such a requirement is prohibitive and perhaps not even
feasible to implement. Therefore, a discrete Monte Carlo solution
method is best suited for situations in which the operator is readily,
if not naturally, formed. Modern nonlinear methods meet this
requirement with Newton methods used in conjunction with Krylov
methods operating on a fully formed matrix for a robust solution
strategy. Furthermore, modern techniques exist that permit the
automatic construction of the Jacobian operator generated within a
Newton method based on the nonlinear residual evaluations, providing
all of the components necessary for a Monte Carlo solver to provide
value. We therefore devise a new nonlinear method in this work based
on the Monte Carlo Synthetic Acceleration algorithm and Newton's
method.

\subsection{Domain Decomposed Monte Carlo Synthetic Acceleration}
\label{subsec:parallel_motivation}

The computational resources required to solve the neutronics and fluid
dynamics problems are tremendous. Recent work in modeling coupled
fluid flow and solid material heat and mass transfer in a reactor
subsystem, similar to the same components of the departure from
nucleate boiling example above, was performed as part of analysis of
the Department of Energy's Consortium for Advanced Simulation of Light
Water Reactors (CASL) modeling and simulation hub. CASL used the
Drekar multiphysics code developed at Sandia National Laboratories
\citep{pawlowski_drekar_2012} for modeling goals in
grid-to-rod-fretting analysis and will use a similar coupled physics
structure for future departure from nucleate boiling analysis with
comparison to experimental data. Using Drekar, multiphysics
simulations have been performed with fully consistent methods for the
solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
leveraging $O(100,000)$ cores on leadership class machines. Neutronics
components to be implemented in CASL for multiphysics analysis, such
as the Exnihilo radiation transport suite developed at Oak Ridge
National Laboratory \citep{evans_denovo:_2010}, compute trillions of
unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
meshes and $O(100,000)$ cores as well. Given the large scale and
complexity of these problems, if we aim to advance multiphysics
solution techniques, then we are motivated to advance the solution of
complex physics problems exploiting leadership class levels of
parallelism.

For Monte Carlo Synthetic Acceleration methods to be viable at the
production scale for nuclear engineering applications, scalable
parallel implementations of the algorithm are required. Reviewing the
literature, Monte Carlo Synthetic Acceleration has yet to be
parallelized and the Neumann-Ulam Monte Carlo method on which it is
built has only been parallelized through history-level parallelism
with full domain replication \citep{alexandrov_efficient_1998}. In
order to solve large linear transport systems with Monte Carlo
Synthetic Acceleration, a domain-decomposed parallel strategy is
required. In the formulation of a parallel Monte Carlo Synthetic
Acceleration algorithm, we recognize that the algorithm occurs in two
stages, an outer iteration performing fixed point iteration and
applying the correction, and an inner Monte Carlo solver that is
generating the correction via the adjoint or forward methods. The
parallel aspects of both these components must be
considered. Therefore, we will develop and implement parallel
algorithms for the fundmental random walk sequence and Monte Carlo
Synthetic Acceleration methods leveraging both the knowledge gained
from the general parallel implementations of subspace methods reviewed
in Appendix~\ref{ch:linear_problem} and modern parallel strategies for
domain decomposed Monte Carlo as developed by the reactor physics
community.

@@

%%---------------------------------------------------------------------------%%
\section{Research and Development Outline}
\label{sec:research_outline}
For some time, the particle transport community has been utilizing
Monte Carlo methods for the solution of transport problems
\citep{lewis_computational_1993}. The partial differential equation
(PDE) community has focused on various deterministic methods for
solutions to linear problems \citep{saad_iterative_2003,
  kelley_iterative_1995}. In between these two areas are a not widely
known group of Monte Carlo methods for solving sparse linear systems
\citep{forsythe_matrix_1950, hammersley_monte_1964,
  halton_sequential_1962, halton_sequential_1994}. In recent years,
these methods have been further developed for radiation transport
problems in the form of Monte Carlo Synthetic Acceleration (MCSA)
\citep{evans_monte_2009, evans_monte_2012} but have yet to be applied
to more general sparse linear systems commonly generated by the
computational physics community. Compared to other methods in this
regime, MCSA offers three attractive qualities; (1) the physics
operator need not be symmetric or positive-definite, thereby reducing
preconditioning complexity, (2) the stochastic nature of the solution
method provides a natural solution to the issue of resiliency, and (3)
is amenable to parallelization using modern methods developed by the
transport community \citep{wagner_hybrid_2010}. The development of
MCSA as a general linear solver and the development of a parallel MCSA
method are new and unique features of this work, providing a framework
with which other issues such as resiliency may be addressed in the
future.

@@

The Forward-Automated Newton-MCSA (FANM) method is developed as new
nonlinear solution method. The key features of FANM are: full Jacobian
generation using modern Forward Automated Differentiation (FAD)
methods \citep{bartlett_automatic_2006}, and MCSA as the inner linear
solver. This method has several attractive properties. First, the
first-order approximation to the Jacobian used in JFNK type methods is
eliminated by generating the Jacobian explicitly with the model
equations through FAD. Second, the Jacobian need not be explicitly
formed by the user but is instead automated through FAD; this
eliminates the complexity of hand-coding derivatives and has also been
demonstrated to be more efficient computationally than evaluating
difference derivatives. Third, unlike GMRES, MCSA does not build a
subspace during iterations. Although the Jacobian must be explicitly
formed to use MCSA, for problems that take more than a few GMRES
iterations to converge the size of the Krylov subspace will grow
beyond that of the Jacobian. Finally, using MCSA to solve the linear
model provides its benefits for preconditioning, potential resiliency,
and parallelism.

%%---------------------------------------------------------------------------%%
\section{Statement of Work}
\label{sec:statement_of_work}

This work will improve the iterative performance and parallel
scalability of solutions to discrete linear and nonlinear transport
problems by researching and developing a new set of domain decomposed
Monte Carlo Synthetic Acceleration methods. Improvements in
performance and scalability will be demonstrated in a wide variety of
numerical experiments using both simple model neutron transport
problems and neutron transport and fluid flow problems generated by
production physics codes.

%%---------------------------------------------------------------------------%%
\section{Outline of the Document}
\label{sec:doc_outline}

To present these new developments this document is arranged in the
following manner. First, in Chapter~\ref{ch:stochastic_methods}, the
fundamentals of the Monte Carlo method for linear systems are
presented. Using this background on Monte Carlo, synthetic
acceleration methods are then presented and new analysis is provided
using a simple model transport problem. Next, in
Chapter~\ref{ch:spn_equations}, the Monte Carlo methods developed in
Chapter~\ref{ch:stochastic_methods} are applied to the neutron
transport problem. Specifically, the $SP_N$ form of the Boltzmann
neutron transport equation is developed and solved using the Monte
Carlo methods. Using a difficult fission reactor fuel assembly
criticality calculation to drive research and development for these
Monte Carlo methods, several important issues regarding Monte Carlo
solver applicability, performance, and preconditioning will be
discussed along with solutions to these problems. Using these
solutions, the Monte Carlo methods are then verified against modern
solution techniques for the $SP_N$ equations in order verify
correctness for the fuel assembly criticality calculation. In
Chapter~\ref{ch:nonlinear_problem}, the Monte Carlo methods are
applied to the Navier-Stokes equations as a model nonlinear fluid flow
system. A new nonlinear solution technique is developed based on
Newton's method and both verified against and compared to conventional
solution techniques. In Chapter~\ref{ch:parallel_methods}, the Monte
Carlo methods are parallelized using modern reactor physics techniques
for domain decomposed particle transport with a full parallel scaling
analysis provided using a leadership-class computing
facility. Finally, the work is summarized in
Chapter~\ref{ch:conclusion} and topics of future work derived from the
results of this research presented.

### 2013-07-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
working on refactoring the introduction chapter

\chapter{Monte Carlo Synthetic Acceleration Methods for the
  Navier-Stokes \\Equations\ }
\label{ch:nonlinear_problem}
Nonlinear transport problems are a common occurrence in single and
multiple physics problems in nuclear engineering. Systems of partial
differential equations, such as the Navier-Stokes equations that
describe fluid flow, yield discrete sets of stiff equations with
nonlinearities present in the variables when discretized by
conventional methods. These sets of equations characterize momentum
and energy transport through diffusion, viscous forces and convection,
and inertial forces. Traditionally, such systems have been solved by
linearizing them in a form where the nonlinearities in the variables
are eliminated and more traditional linear methods can be used for
solutions. Often characterized as segregated methods where physics
operators are split and their action on the system approximated in
steps, such methods lack consistency and accuracy in resolving the
nonlinear component of the solution. In the last 30 years, fully
consistent nonlinear methods based on Newton's method have become more
popular and many advances have been made in the nuclear engineering
field to employ these methods.

In the context of solving standalone linear systems, Monte Carlo
methods do not provide significant merit over subspace methods due to
the fact that the linear operator must be explicitly formed. For many
applications including higher fidelity neutron transport
discretizations beyond the $SP_N$ equations solved in the previous
chapter, such a requirement is prohibitive and perhaps not even
feasible to implement. Therefore, a discrete Monte Carlo solution
method is best suited for situations in which the operator is readily,
if not naturally, formed. Modern nonlinear methods meet this
requirement with Newton methods used in conjunction with Krylov
methods operating on a fully formed matrix for a robust solution
strategy. Furthermore, modern techniques exist that permit the
automatic construction of the Jacobian operator generated within a
Newton method based on the nonlinear residual evaluations, providing
all of the components necessary for a Monte Carlo solver to provide
value. We therefore devise a new nonlinear method based on the MCSA
algorithm and Newton's method.

In this chapter the fundamentals of Newton's method will be presented
along with a brief discussion of inexact Newton methods. Newton-Krylov
methods, a variant of inexact Newton methods, will be outlined
including with two common techniques for numerically representing the
Jacobian most commonly used in today's applications. Following this,
the new algorithm will be presented and implementation details will be
discussed. Using the new algorithm and a production fluid physics
code, a group of challenging nonlinear benchmark problems for the
Navier-Stokes equations will be used to verify the correctness of the
method relative to solutions obtained using conventional methods in
various flow regimes and geometries. Finally, we will compare the
performance of the new method against contemporary Newton-Krylov
methods for the same benchmark problems.

>>>>

\chapter{Monte Carlo Synthetic Acceleration Methods for the
  Navier-Stokes \\Equations\ }
\label{ch:nonlinear_problem}

In this chapter the fundamentals of Newton's method will be presented
along with a brief discussion of inexact Newton methods. Newton-Krylov
methods, a variant of inexact Newton methods, will be outlined
including with two common techniques for numerically representing the
Jacobian most commonly used in today's applications. Following this, a
new algorithm utilizing Monte Carlo Synthetic Acceleration will be
presented and implementation details will be discussed. Using the new
algorithm and a production fluid physics code, a group of challenging
nonlinear benchmark problems for the Navier-Stokes equations will be
used to verify the correctness of the method relative to solutions
obtained using conventional methods in various flow regimes and
geometries. Finally, we will compare the performance of the new method
against contemporary Newton-Krylov methods for the same benchmark
problems.

### 2013-07-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
working on refactoring the introduction chapter

\chapter{Parallel Monte Carlo Synthetic \\ Acceleration Methods\ }
\label{ch:parallel_methods}
For MCSA and FANM to be viable at the production scale for nuclear
engineering applications, scalable parallel implementations of the
algorithm are required. Reviewing the literature, MCSA has yet to be
parallelized and the Neumann-Ulam Monte Carlo method on which it is
built has only been parallelized through history-level parallelism
with full domain replication \citep{alexandrov_efficient_1998}. In
order to solve large linear transport systems with MCSA, a
domain-decomposed parallel strategy is required. In the formulation of
a parallel MCSA algorithm, we recognize that the algorithm occurs in
two stages, an outer iteration performing fixed point iteration and
applying the correction, and an inner Monte Carlo solver that is
generating the correction via the adjoint or forward methods. The
parallel aspects of both these components must be
considered. Therefore, we will develop and implement parallel
algorithms for the Neumann-Ulam and MCSA methods leveraging both the
knowledge gained from the general parallel implementations of Krylov
methods reviewed in Appendix~\ref{ch:linear_problem} and modern
parallel strategies for domain decomposed Monte Carlo as developed by
the reactor physics community.

>>>>

\chapter{Parallel Monte Carlo Synthetic \\ Acceleration Methods\ }
\label{ch:parallel_methods}
